BIOST 578 HW 3
========================================================
 
```{r, message=FALSE, echo=FALSE}
## load libraries
library(GEOquery)
library(limma)
library(pheatmap)
library(GSEABase)
```

## Get data from GEO 

```{r}
#gd <- getGEO("GSE45735", destdir = "./Data/")
gd <- getGEO(filename = "./Data/GSE45735_series_matrix.txt")  
pd <- pData(gd)

## the T14 file is problematic and needs to be fixed by hand
#getGEOSuppFiles("GSE45735", makeDirectory=FALSE, baseDir = "./Data/")  ## need to manually unzip .gz file first before proceeding

## note the regular expression to grep file names
files <- list.files(path = "./Data/", pattern = "GSE45735_T.*.txt", full.names = TRUE) 
file_list <- lapply(files, read.table, header=TRUE)

## remove duplicated rows
file_list_unique <- lapply(file_list, function(x){x<-x[!duplicated(x$Gene),]; 
                                                  x <- x[order(x$Gene),]; 
                                                  rownames(x) <- x$Gene;
                                                  x[,-1]})

## take the intersection of all genes
gene_list <- Reduce(intersect, lapply(file_list_unique, rownames))
file_list_unique <- lapply(file_list_unique, "[", gene_list,)
matrix <- as.matrix(do.call(cbind, file_list_unique))

## clean up the pData
pd_small <- pd[!grepl("T13_Day8",pd$title),]
pd_small$Day <- sapply(strsplit(gsub(" \\[PBMC\\]", "", pd_small$title),"_"),"[",2)
pd_small$subject <- sapply(strsplit(gsub(" \\[PBMC\\]", "", pd_small$title),"_"),"[",1)
colnames(matrix) <- rownames(pd_small)
```

## Use voom and limma to find genes differentially expressed at each time point compared to baseline (day 0). Use an FDR cutoff of 0.01. Display your results using pheatmap showing the log fold-change of differentially expressed genes grouped by time point.

```{r}
## add one to count
new_set <- ExpressionSet(assayData = matrix+1)
pData(new_set) <- pd_small

## voom + limma
design <- model.matrix(~subject+Day, new_set)
new_set_voom <- voom(new_set, design = design)
lm <- lmFit(new_set_voom, design)
eb <- eBayes(lm)

## get adjusted P value for each coefficient associated with Day for each probe
res.genes <- vector("list", length = 10)
for(i in 1:10) {
  coef <- paste0("DayDay", i)
  res.genes[[i]] <- topTable(eb, coef = coef, number=Inf)
}

## visualize results
PValue <- sapply(res.genes, function(x){x$adj.P.Val})
LogFC <- sapply(res.genes, function(x){x$logFC})
rownames(LogFC) <- rownames(res.genes[[1]])
PValue_min <- rowMin(PValue)
LogFC_small <- LogFC[PValue_min<0.01, ]  # FDR cutoff of 0.01
anno <- data.frame(Time=paste0("Day", 1:10))
rownames(anno) <- colnames(LogFC_small) <- paste0("Day", 1:10)
pheatmap(LogFC_small, cluster_cols=FALSE)
```


## Perform a GSEA analysis using camera and the MSigDB Reactome pathway gene signatures. Display your results using pheatmap, again group by timepoint.

```{r}
## convert gene sets to gene indices
c2_set <- getGmt("./Data/c2.all.v4.0.symbols.gmt")
gene_ids <- geneIds(c2_set)
sets_indices <- symbols2indices(gene_ids, rownames(new_set))

## find enriched gene sets over time
res.gsea <- vector("list",length = 10)
for(i in 1:10)
  {
  contrast <- paste0("DayDay", i)
  cont_matrix <- makeContrasts(contrast, levels=design)
  res.gsea[[i]] <- camera(new_set_voom, sets_indices, design=design, contrast=cont_matrix, sort=FALSE)
}

## visualize results
PValueGSEA <- sapply(res.gsea, function(x){ifelse(x$Direction=="Up", -10*log10(x$PValue), 10*log10(x$PValue))})
rownames(PValueGSEA) <- rownames(res.gsea[[1]])
PValueGSEA_max <- rowMax(abs(PValueGSEA))
PValueGSEA_small <- PValueGSEA[PValueGSEA_max>30, ]
anno <- data.frame(Time=paste0("Day",1:10))
rownames(anno) <- colnames(PValueGSEA_small)  <- paste0("Day",1:10)
pheatmap(PValueGSEA_small, cluster_cols=FALSE)
```